---
title: "The Book of R Chapter 3 Exercises"
Subject: Matrices and Arrays
output: html_notebook
---

**Exercise 3.1**

**a.** Construct and store a 4 X 2 matrix that's filled row-wise with the values 4.3, 3.1, 8.2, 3.2, 0.9, 1.6, and 6.5. in that order.

```{r}
a = matrix(data = c(4.3, 3.1, 8.2, 8.2, 3.2, 0.9, 1.6, 6.5), nrow = 4, ncol =2, byrow = TRUE)
a
```

**b.** Confirm the dimensions of the matrix from (a) are 3 x 2 if you remove any one row.

```{r}
b <- a[-1,] #Deletes the first row of matrix a
dim(b)      #Gives the dimension of matrix b
b
```

The matrix b has dimensions 3 rows x 2 columns (line 1 above).

**c.** Overwrite the second column of the matrix from (a) with that same column sorted from smallest to largest.

```{r}
a[, 2] <- sort(a[, 2], decreasing = FALSE)
a
```

**d.** What does R return if you delete the fourth row and the first column from (c)? Use matrix to ensure the result is a single-column matrix, rather than a vector.

```{r}
d <- matrix(data = a[-4, -1])
d
```

**e.** Store the bottom four elements of (c) as a new 2 x 2 matrix.

```{r}
e <- a[-(1:2),]
e

```

**f.** Overwrite, in this order, the elements of (c) at positions (4, 2), (1, 2), (4, 1), and (1, 1) with -1/2 of the two values on the diagonal of (e).

```{r}
a[c(4, 1), 2:1] <- -0.5* diag(x = e)
a
```

The above first filled (4, 2) with -1.6 (-0.5 \* 3.2, which is -0.5\*e[1,1]),

second it filled (1, 2) with -4.1 (-0.5 \* 8.2, which is -0.5 \* e[2, 2]),

third it filled (4, 1) with -1.6,

and last,

it filled (1, 1) with -4.1

**Exercise 3.2**

**a.** Calculate the following: (matrix multiplication and subtraction on p. 52 of Davies.

```{r}
A <- cbind(c(1, 2, 7), c(2, 4, 6))
B <- cbind(c(10, 30, 50), c(20, 40, 60))
A
B
a <- (2/7)*(A - B)
a

```

**b.** Store these two matrices (p. 52 of Davies).

```{r}
A <- matrix(c(1, 2, 7))
B <- matrix(c(3, 4, 8))
A
B
```

Which of the following multiplications are possible? For those that are , compute the result.

i.  A \* B

    This is not possible because the matrices are the same dimension. Matrix A must have the same number of columns as matrix B has rows.

```{r}
dim(A)
dim(B)
```

ii. A^⊤^ \* B

This is possible because A^⊤^ has 3 columns and B has three rows.

```{r}
ii <- t(A)
dim(ii)
dim(B)
ii %*% B
```

iii. B^⊤^ \* (A \* A^⊤^)

This is possible because B^⊤^ has 1 column and (A \* A^⊤^) has one row.

```{r}
dim(t(B))
dim(A %*% t(A))
iii <- t(B) %*% (A %*% t(A))
iii
```

iv. (A \* A^⊤^) \* B^⊤^

This is not possible because (A \* A^⊤^) has three columns and B^⊤^ has one row.

```{r}
dim(A %*% t(A))
dim(t(B))
```

v.  [(B \* B^⊤^) + (A \* A^⊤^) - 100*I*~3~ ]^-1^

This is possible because (B \* B^⊤^), (A \* A^⊤^), and *I*~3~ are all 3 x 3 matrices.

```{r}
vB <- B %*% t(B)
vA <- A %*% t(A)
I <- diag(x = 3)
(vB + vA - (100 * I))^(-1)

```

**c.** For (matrix given in Davies) confirm that A^-1^ \* A - I~4~ provides a 4 x 4 matrix of zeros.

```{r}
A <- cbind(c(2, 0, 0, 0), c(0, 3, 0, 0), c(0, 0, 5, 0), c(0, 0, 0, -1))
Ainv <- solve(A)
I <- diag(x = 4)
A
Ainv
I
Ainv %*% A - I
```

**Exercise 3.3**

**a.** Create and store a three-dimensional array with six layers of a 4 x 2 matrix, filled with a decreasing sequence of values between 4.8 and 0.1 of the appropriate length.\
The array will be 4 rows x 2 columns x 6 layers so it will take 48 values that decrease in 0.1 increments from 4.8.

```{r}
a <- array(data = seq(from = 4.8, to = 0.1, length.out = 48), dim = c(4, 2, 6))
a
```

**b.** Extract and store as a new object the fourth- and first-row elements, in that order, of the second column only of all layers of (a).

Note that these will be combined as rows in the resulting matrix.

```{r}
b <- a[c(4, 1), 2, ]
b
```

**c.** Use a fourfold repetition of the second row of the matrix formed in (b) to fill a new array of dimensions 2 x 2 x 2 x 3.

```{r}
c <- array(data = rep(b[2,], times = 4), dim = c(2, 2, 2, 3))
c

```

**d.** Create a new array comprised of the results of deleting the sixth layer of (a).

```{r}
d <- a[ , , -6]
d
```

**e.** Overwrite the second and fourth row elements of the second column of layers 1,3, and 5 of (d) with -99.

```{r}
d[c(2, 4), 2, c(1, 3, 5)] <- (-99)
d
```
